---
title: "R Workshop - Part 5 (Joins, Strings and Mapping)"
author: "Peter Herman"
date: "2023-06-26"
output: slidy_presentation
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(readr)
library(dplyr)
library(tidyr)
library(sf)
library(stringr)

setwd(r"()")
raw_school_data <- read_delim("Example_Datasets/California_Academic_Performance_Index.txt",delim="\t")
raw_census_shapefiles <- st_read("Example_Datasets/California_Counties.geojson")

census_shapefile_with_dummy_row <- raw_census_shapefiles %>%
  mutate(dummy = 1)

options(warn = -1)
options(datatable.verbose = FALSE)

```

## Reading in Data

Press K in your browser to avoid the site showing the next slide automatically.

As we did with the last exercise, the typical R script opens with importing libraries, setting the working directory, and reading in data. 

```{r E1, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5, echo=FALSE,message=FALSE}

library(dplyr)
library(stringr)
library(plotly)
library(tidycensus)

setwd(r"()")

raw_school_data <- read_delim("Example_Datasets/California_Academic_Performance_Index.txt",delim="\t")

print(raw_school_data %>% head())

```

## Reading in Spatial Data

For manipulating spatial/GIS data, we use the st_read command from the sf library.

Objects with class sf are just like data frames, but with a geometry column that includes a geometry type (multipolgyon in this case) and coordinates.

```{r E2, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5,message=FALSE}

library(sf)

raw_census_shapefiles <- st_read("Example_Datasets/California_Counties.geojson")

class(raw_census_shapefiles)

print(head(raw_census_shapefiles))

```

## Manipulating Spatial Data

We can manipulate an sf object using dplyr commands. Here we're adding a column called dummy that has a value of 1 for all rows. 

```{r E3, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5,message=FALSE}

census_shapefile_with_dummy_row <- raw_census_shapefiles %>%
  mutate(dummy = 1)

print(head(census_shapefile_with_dummy_row))

```


## Manipulating a string variable to perform a data join

The two data sets we want to join do not have any fields in common. The closest we have are the county name fields on both files

```{r E4-setup}

census_shapefile_with_dummy_row <- raw_census_shapefiles %>%
  mutate(dummy = 1)

```

```{r E4, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5,message=FALSE}

# The counties in the Census SF object have names that include county and state
census_shapefile_with_dummy_row %>% arrange(NAME) %>% pull(NAME) %>% head() 

# The counties in the school dataset do not 
raw_school_data %>% arrange(cname)  %>% pull(cname) %>% unique() %>% head()

```

## Joining two datasets

We want to compare the median academic performance index score among high schools in each county to the median household income in each county.

We have two options here, the easy way is to use the str_c function from the stringr library to concatenate the county names in the school data with ", County, California"

```{r E5-setup}

census_shapefile_with_dummy_row <- raw_census_shapefiles %>%
  mutate(dummy = 1)

```

```{r E5, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5,message=FALSE}

# First, we want to have one row per county in our data frame
median_api_per_hs_in_county <- raw_school_data %>%
  filter(stype == "H") %>%
  group_by(cname) %>%
  summarize(median_api00 = median(api00)) %>%
  ungroup()

print(head(median_api_per_hs_in_county))
```

## Joining two datasets

We want to compare the median academic performance index score among high schools in each county to the median household income in each county.

We have two options here, the easy way is to use the str_c function from the stringr library to concatenate the county names in the school data with ", County, California"

```{r E5_v12-setup}

census_shapefile_with_dummy_row <- raw_census_shapefiles %>%
  mutate(dummy = 1)

# First, we want to have one row per county in our data frame
median_api_per_hs_in_county <- raw_school_data %>%
  filter(stype == "H") %>%
  group_by(cname) %>%
  summarize(median_api00 = median(api00)) %>%
  ungroup()

```

```{r E5_v12, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5,message=FALSE}
# Add the field we will be using for our data join
school_data_to_join <- median_api_per_hs_in_county %>%
  mutate(county_name_with_state = str_c(cname," County, California"))

# Use the left_join function from dplyr to do the join
# the join_by statement has an expression that is TRUE for every row we want to join 
school_data_with_income <- school_data_to_join %>%
  left_join(census_shapefile_with_dummy_row,by=join_by("county_name_with_state" == "NAME"))

# We can test if our data joined by summing up the dummy variable and testing if it's equal to the number of rows in our dataset 
nrow(school_data_with_income)
print(sum(school_data_with_income %>% pull(dummy),na.rm=TRUE))

```

```{r E5_v2-setup}

# First, we want to have one row per county in our data frame
median_api_per_hs_in_county <- raw_school_data %>%
  filter(stype == "H") %>%
  group_by(cname) %>%
  summarize(median_api00 = median(api00)) %>%
  ungroup()

# Add the field we will be using for our data join
school_data_to_join <- median_api_per_hs_in_county %>%
  mutate(county_name_with_state = str_c(cname," County, California"))

# Use the left_join function from dplyr to do the join
# the join_by statement has an expression that is TRUE for every row we want to join 
school_data_with_income <- school_data_to_join %>%
  left_join(census_shapefile_with_dummy_row,by=join_by("county_name_with_state" == "NAME")) 

```

## Finding rows that did not join

We can use dplyr filter with our dummy variable to find the rows that did not join. It appears as though San Francisco is missing a space in the school data frame

```{r E5_v2, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5,message=FALSE}

# There is row that did not match
print(nrow(school_data_with_income)) 
print(sum(school_data_with_income %>% pull(dummy),na.rm=TRUE))

# Filter to the rows that did not match
Unmatched_Rows <- school_data_with_income %>%
  filter(is.na(dummy))

# Pull out the names of the counties that did not match
Unmatched_Rows %>% pull(county_name_with_state) %>% unique()

```

## Using case_when to fix our data join

We can use the case_when function from dplyr, fix the one county name that did not match, and leave the rest as is. We can use .default to avoid writing a conditional for every case. 
```{r E5_v3-setup}

median_api_per_hs_in_county <- raw_school_data %>%
  filter(stype == "H") %>%
  group_by(cname) %>%
  summarize(median_api00 = median(api00)) %>%
  ungroup()

```

```{r E5_v3, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5,message=FALSE}

# Add the field we will be using for our data join, using case_when for the county name that did not match
school_data_to_join <- median_api_per_hs_in_county %>%
  mutate(county_name_with_state = case_when(
    cname == "SanFrancisco"  ~ "San Francisco County, California",
    .default = str_c(cname," County, California"))
  ) 

# Do a left join with the corrected field
school_data_with_income <- school_data_to_join %>%
  left_join(census_shapefile_with_dummy_row,by=join_by("county_name_with_state" == "NAME"))

# Test if all fields matched
print(nrow(school_data_with_income) == sum(school_data_with_income %>% pull(dummy)))

```


## Other string manipulation options - stringr

Alternatively, we can use a number of different functions in the stringr package to remove the ", County, California" from the census data, rather than add it to the school data

```{r E7, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5,message=FALSE}

# We could use the str_remove function to identify the part of the string we want to remove
# or use str_repalce to replace the part of the string we want to remove with an empty string
# Or str_extract to just extract the part of the string we want with a regular expression
census_shapefile_with_cleaned_names <- census_shapefile_with_dummy_row %>%
  mutate(
    Clean_Conunty_Name1 = str_remove(NAME," County, California"),
    Clean_County_Name2 = str_replace(NAME," County, California",""),
    Clean_County_Name3 = str_extract(NAME,  "([A-z]+| )(?= County)")
  )

# Taking a look at our results
census_shapefile_with_cleaned_names %>%
  select(Clean_Conunty_Name1,Clean_County_Name2,Clean_County_Name3) %>%
  head() %>%
  print()


```

## Other string manipulation options - separate

Another alternative to extracting the county name from the census data frame is to use the separate function from dplyr

```{r E8, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5,message=FALSE}

# Separate takes one column, a list of new column names, and a separator
census_shapefile_with_cleaned_names <- census_shapefile_with_dummy_row %>%
  separate(NAME, c("COUNTY_NAME","STATE_NAME"), sep = ' County,')


# Taking a look at our results
census_shapefile_with_cleaned_names %>%
  select(COUNTY_NAME,STATE_NAME) %>%
  head() %>%
  print()


```

## Mapping our data

Now that we have our median api score appended to a sf object with a geometry column, we can map our data using the plotly library. 

```{r E9-setup}

# Add the field we will be using for our data join, using case_when for the county name that did not match
median_api_per_hs_in_county <- raw_school_data %>%
  filter(stype == "H") %>%
  group_by(cname) %>%
  summarize(median_api00 = median(api00)) %>%
  ungroup()
# 
school_data_to_join <- median_api_per_hs_in_county %>%
  mutate(county_name_with_state = case_when(
    cname == "SanFrancisco"  ~ "San Francisco County, California",
    .default = str_c(cname," County, California"))
  )

options(datatable.verbose = FALSE)

```

```{r E9, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5,message=FALSE}

library(plotly)

# Repeat our left join- but add a command to keep the resulting dataset as an sf object 
school_data_with_income <- census_shapefile_with_dummy_row %>%
  left_join(school_data_to_join,by=join_by("NAME" == "county_name_with_state")) %>%
  st_sf()

# Map our data
api_score00_map <-plot_ly(
    school_data_with_income, 
    split = ~cname,
    color = ~median_api00,
    alpha = 1,
    showlegend = FALSE
)

api_score00_map

```

```{r E10-setup}

# Add the field we will be using for our data join, using case_when for the county name that did not match
median_api_per_hs_in_county <- raw_school_data %>%
  filter(stype == "H") %>%
  group_by(cname) %>%
  summarize(median_api00 = median(api00)) %>%
  ungroup()
# 
school_data_to_join <- median_api_per_hs_in_county %>%
  mutate(county_name_with_state = case_when(
    cname == "SanFrancisco"  ~ "San Francisco County, California",
    .default = str_c(cname," County, California"))
  )

school_data_with_income <- census_shapefile_with_dummy_row %>%
  left_join(school_data_to_join,by=join_by("NAME" == "county_name_with_state")) %>%
  st_sf() %>%
  st_transform(st_crs(4326))

options(datatable.verbose = FALSE)

```

## Mapping our data - visualization options

There are many, many options available for improving the readability of our plotly maps. Here we have changed the colors in the map to a green color scale, and added a black border.

```{r E10, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5,message=FALSE}


# Map our data
api_score00_map_custom <-plot_ly(
    school_data_with_income, 
    split = ~cname,
    color = ~median_api00,
    alpha = 1,
    showlegend = FALSE,
    stroke= I("black"),
    colors = "Greens"
)   

api_score00_map_custom

```


