---
title: "R Workshop - Part 3 (Transforming Data)"
author: "Peter Herman"
date: "2023-06-26"
output: slidy_presentation
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(readr)
library(dplyr)
library(tidyr)

setwd(r"()")
raw_school_data <- read_delim("Example_Datasets/California_Academic_Performance_Index.txt",delim="\t")

```

## Reading in Data

Press K in your browser to avoid the site showing the next slide automatically.

As we did with the last exercise, the typical R script opens with importing libraries, setting the working directory, and reading in data

```{r C1, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

library(dplyr)

setwd(r"()")

raw_school_data <- read_delim("Example_Datasets/California_Academic_Performance_Index.txt",delim="\t")



```

## Transforming Data - Filtering

We can use the filter() function in dplyr to filter our data frame

```{r C2, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

library(dplyr)
print(table(raw_school_data$stype))
print(dim(raw_school_data))

# In R, the == operator means, test for equivalence
print(1 == 1)
print(1 == 2)

# We can use the == operator in the filter function to filter our data set
high_schools_only <- raw_school_data %>%
  filter(stype == "H")

print(table(high_schools_only$stype))
print(dim(high_schools_only))

```

## Transforming Data - Filtering with operators

All three of the filter functions below yield the same data frame

```{r C3, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

print(table(raw_school_data$stype))

# We can use the or operator | in the filter function to filter to either high schools or middle schools
high_schools_or_middle_schools1 <- raw_school_data %>%
  filter(stype == "H" | stype == "M")

# Alternatively we can use the NOT operator !  to filter every row that is not an elementary school
high_schools_or_middle_schools2 <- raw_school_data %>%
  filter(!stype == "E")

# We could also use the %in% operator to create a vector of acceptable values
high_schools_or_middle_schools3 <- raw_school_data %>%
  filter(stype == c("H", "M"))


print(table(high_schools_or_middle_schools1$stype))
print(table(high_schools_or_middle_schools2$stype))
print(table(high_schools_or_middle_schools3$stype))


```

## Transforming Data - Sort

The arrange function will sort the data frame

```{r C4, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# For demo purposes, let's take the first five rows of the data frame using head()
# Sorting by school name 
Sorted_by_Name <- raw_school_data %>%
  arrange(sname) %>%
  head()

print(Sorted_by_Name)

# You can sort on multiple columns by adding multiple column names to the arrange function
Sorted_by_Name_and_School_Type <- raw_school_data %>%
  arrange(stype,sname) %>%
  head()

print(Sorted_by_Name_and_School_Type)


```

## Transforming Data - renaming columns

The rename function is used to rename columns. 
Within dplyr functions, use = for assignment.

```{r C5, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# See the first five rows of the data frame
print(raw_school_data %>% head())

# Rename columns
# We can use the back quote to give R field names that have spaces
Data_with_Renamed_Columns <- raw_school_data %>%
  rename(
    `School Type` = stype,
     school_name = sname
  )

print(Data_with_Renamed_Columns %>% head())


```

## Transforming Data - removing and rearranging columns

We can use the select function to subset our data or rearrange it

```{r C6, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# See the first five rows of the data frame
print(raw_school_data %>% head())

# We put the columns we want to keep into a vector within the select statement
First_three_columns <- raw_school_data %>%
  select(c(school_id,stype))

print(First_three_columns %>% head())

```

We can also put the columns in a different order using select

```{r C6_2, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

First_three_columns_reordered <- raw_school_data %>%
  select(c(name,school_id))

print(First_three_columns_reordered %>% head())

```

We can also remove columns using select with the minus sign - 

```{r C6_3, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

First_column_only <- raw_school_data %>%
  select(c(name,school_id,stype)) %>%
  select(-c(stype,school_id))

print(First_column_only %>% head())

```

## Transforming Data - adding columns

The mutate function is used to add columns

```{r C7, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# Our data has two years of test score data (the test is called API)
Test_Columns_Only <- raw_school_data %>%
  select(c(school_id,name,api00,api99)) 

print(Test_Columns_Only %>% head())

# We can use mutate to create a new field with difference in test score between 2000 and 1999
Test_Scores_with_differences <- Test_Columns_Only %>%
  mutate(Test_Score_Difference = api99 - api00)

print(Test_Scores_with_differences %>% head())

```

## Transforming Data - adding multiple columns

We can create multiple fields within one call to mutate()

```{r C8, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# Our data has two years of test score data (the test is called API)
Test_Score_Data <- raw_school_data %>%
  select(c(school_id,name,api00,api99)) %>%
  mutate(Test_Score_Difference = api00 - api99,
         Raw_Change_Rate = Test_Score_Difference/api99,
         Change_Percent = round(Raw_Change_Rate*100,digits = 1),
         Change_Percent_Character = paste0(as.character(Change_Percent),"%")
         ) %>%
  select(name,Test_Score_Difference,Raw_Change_Rate,Change_Percent,Change_Percent_Character)

print(Test_Score_Data %>% head())

```

## Transforming Data - Recoding

We can recode variables using the case_when function within mutate()
```{r C9, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# Our data has two years of test score data (the test is called API)
Test_Score_Data <- raw_school_data %>%
  select(c(school_id,name,api00,api99)) %>%
  mutate(Test_Score_Difference = api00 - api99) %>%
  mutate(Difference_Type= case_when(
   Test_Score_Difference > 0 ~ "Positive",
   Test_Score_Difference < 0 ~ "Negative"
  ))

print(Test_Score_Data %>% head())

# Notice we have some NA's because we didn't account for if there's no difference in exam score
print(table(Test_Score_Data$Difference_Type,useNA = c("always")))

```

## Transforming Data - transforming multiple variables at once

We can use mutate_if to transform all columns that pass a logic test

```{r C10, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# Start with our normal test score dataset
Test_Score_Data <- raw_school_data %>%
  select(c(school_id,name,api00,api99)) 

# We can change all numeric fields to character using mutate_if
# mutate if takes two arguments, a logic test that runs on each column, and a function to apply 
print(is.numeric(9))
print(is.numeric("apple"))
print(as.character(9))

Character_Test_Score_Data <- Test_Score_Data %>%
  mutate_if(is.numeric, as.character)

print(Character_Test_Score_Data %>% head())

# You can use your own custom functions in a mutate statement
double_my_number <- function(num){return(num*2)}

Doubled_Test_Scores <- Test_Score_Data %>%
  mutate_if(is.numeric, double_my_number)

print(Doubled_Test_Scores %>% head())


```


## Transforming Data - transforming multiple variables at once

We can use mutate_at to transform all columns within a vector

```{r C11, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# Start with our normal test score dataset
Test_Score_Data <- raw_school_data %>%
  select(c(school_id,name,api00,api99)) 

# Using mutate_at we can select the fields we want to transform individually
# This is one of the few times we have to refer to variables in quotes in a dplyr function
Test_Score_Data_Api100_Character <- Test_Score_Data %>%
  select(c(school_id,name,api00,api99)) %>%
  mutate_at(c("api00"),as.character)

# Notice how api00 is now a character variable
print(Test_Score_Data_Api100_Character %>% head())


```

## Transforming Data - removing duplicates

We can use the length and unique functions to see if there are duplicates in our dataset

```{r C12, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# The unique function will create a vector with the unique values of a column
print(raw_school_data %>% pull(stype) %>% unique())

# The length function will tell us the length of a vector
print(raw_school_data %>% pull(school_id) %>% unique() %>% length())
# It appears as though we have one row per school_id
print( dim(raw_school_data))

# However, there are schools with repeated names! 
print(raw_school_data %>% pull(name) %>% unique() %>% length())

# If we wanted one row per school name
Unique_School_Name_Dataset <- raw_school_data %>%
  distinct(name,.keep_all = TRUE)

# The keep all argument allows us to keep all the other columns in our data frame
print(dim(Unique_School_Name_Dataset))

```

## Transforming Data - rotating data

Let's say we wanted our data in "wide" instead of "long" format
For example, if we wanted one column per school rather than one row per school

```{r C13, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

library(tidyr)

# For demo purposes, let's just take the first 5 rows and test score columns
Test_Score_Data <- raw_school_data %>%
  select(c(name,api00,api99)) %>%
  head()

print(Test_Score_Data)

Wide_Formatted_Data <- Test_Score_Data %>%
  select(name,api00) %>%
  pivot_wider(
    names_from = name,
    values_from = api00
  ) 
print(Wide_Formatted_Data)


```

## Transforming Data - appending data frames together

Let's say we were creating a data frame with 2 rows, test scores in 2000 and in 1999, and one column per school

```{r C14, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

library(tidyr)

# For demo purposes, let's just take the first 5 rows and test score columns
Test_Score_Data <- raw_school_data %>%
  select(c(name,api00,api99)) %>%
  head()

# Create two wide versions of the data
Wide_data_api00 <- Test_Score_Data %>%
  select(name,api00) %>%
  pivot_wider(
    names_from = name,
    values_from = api00
  ) %>%
  mutate(Label = "API Score 2000")

Wide_data_api99 <- Test_Score_Data %>%
  select(name,api99) %>%
  pivot_wider(
    names_from = name,
    values_from = api99
  ) %>%
  mutate(Label = "API Score 1999")

# Stack them using bind_rows
Wide_Data_Stacked <- Wide_data_api00 %>%
  bind_rows(Wide_data_api99)

print(Wide_Data_Stacked)

# Move the label column to the front
Wide_Data_Stacked_Reordered <- Wide_Data_Stacked %>%
  relocate(Label)

print(Wide_Data_Stacked_Reordered)

```

## Exporting Data 

We can export using write_delim(), or in a native R format using write_rds()

```{r C15, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# For demo purposes, let's just take the first 5 rows and test score columns
Test_Score_Data <- raw_school_data %>%
  select(c(name,api00,api99)) %>%
  head()

# We can use these two functions to write data from R (does not work in the browser)
# write_delim(Test_Score_Data,"my_output_dataset.csv",delim=",")
# write_rds(Test_Score_Data,"my_output_dataset.rds")


```

