---
title: "R Workshop - Part 4 (Summarizing and Plotting Data)"
author: "Peter Herman"
date: "2023-06-26"
output: slidy_presentation
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(readr)
library(dplyr)
library(tidyr)

setwd(r"()")
raw_school_data <- read_delim("Example_Datasets/California_Academic_Performance_Index.txt",delim="\t")

```

## Reading in Data

Press K in your browser to avoid the site showing the next slide automatically.

As we did with the last exercise, the typical R script opens with importing libraries, setting the working directory, and reading in data. We will be using the plotly library this time.

```{r D1, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

library(dplyr)
library(plotly)

setwd(r"(C:\Users\peter\OneDrive\Documents\Work\Field_Directors)")

raw_school_data <- read_delim("Example_Datasets/California_Academic_Performance_Index.txt",delim="\t")


```

## Summarizing Data

Using the dplyr summarize() function, we can collapse data into a single row.
Summarize is syntactically similar to mutate

```{r D2, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# enroll is the number of students enrolled in a school
enroll_Summary_Stats <- raw_school_data %>% 
  filter(!is.na(enroll)) %>%
  summarize(
    Mean_enroll = mean(enroll),
    Max_enroll = max(enroll),
    Min_enroll = min(enroll)
  )

print(enroll_Summary_Stats)

# There's a special function called n() which gives the count of rows in a summary group
enroll_Summary_Stats2 <- raw_school_data %>% 
  filter(!is.na(enroll)) %>%
  summarize(
    Mean_enroll = mean(enroll),
    Max_enroll = max(enroll),
    Min_enroll = min(enroll),
    `Number of Schools` = n()
  )

print(enroll_Summary_Stats2)
```

## Summarizing Data - By Group

Using the dplyr group_by() function, we can change the unit of analysis to an individual group. For example, school type

```{r D3, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# Recall we have a variable,stype, that states whether each row is a elementary, middle or high school
enroll_Summary_Stats_by_Stype <- raw_school_data %>% 
  filter(!is.na(enroll)) %>%
  group_by(stype) %>%
  summarize(
    Mean_enroll = mean(enroll),
    Max_enroll = max(enroll),
    Min_enroll = min(enroll),
   `Number of Schools` = n()
  )

print(enroll_Summary_Stats_by_Stype)

```

## Plotting Data - Bar Charts

Let's say we wanted to compare the mean scores of each school type individually

```{r D4, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# First, let's use the summary dataset we created on the last slide
enroll_Summary_Stats_by_Stype <- raw_school_data %>% 
  filter(!is.na(enroll)) %>%
  group_by(stype) %>%
  summarize(
    Mean_enroll = mean(enroll),
    Max_enroll = max(enroll),
    Min_enroll = min(enroll),
   `Number of Schools` = n()
  )

# We can create a bar chart using the plot_ly function from the plotly library
# the tilde sign means "evaluate" in R, it's rarely used out of this context
library(plotly)
bar_chart <- plot_ly(data =enroll_Summary_Stats_by_Stype,
                       type = "bar",
                       y = ~stype ,
                       x = ~`Mean_enroll`)

bar_chart 

# We can make a horizontal bar chart by switching the variable in the x and y arguments

```


## Plotting Data - Grouped Bar Charts

Let's say we wanted to compare the mean scores of each school type, as wenroll as the max and min

```{r D5, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# First, let's use the summary dataset we created on the last slide
enroll_Summary_Stats_by_Stype <- raw_school_data %>% 
  filter(!is.na(enroll)) %>%
  group_by(stype) %>%
  summarize(
    Mean_enroll = mean(enroll),
    Max_enroll = max(enroll),
    Min_enroll = min(enroll),
   `Number of Schools` = n()
  )

# Make the min bar chart, then use add_trace to add the other types of variables
# We do min first, as we want the order to be min, mean, max
min_bar_chart <- plot_ly(data =enroll_Summary_Stats_by_Stype,
                       type = "bar",
                       x = ~stype ,
                       y = ~`Min_enroll`,
                       name = "Min Enrollment")

min_bar_chart_with_mean <- min_bar_chart %>%
  add_trace(y = ~`Mean_enroll`,name = "Mean Enrollment")

min_bar_chart_with_mean_and_max <- min_bar_chart_with_mean %>%
  add_trace(y = ~`Max_enroll`,name = "Max Enrollment")

min_bar_chart_with_mean_and_max 


```


## Plotting Data - Changing the Axis Titles

We can change the appearance of the chart using the layout function. There are many many things one can adjust using this function: https://plotly.com/r/reference/#Layout_and_layout_style_objects

```{r D6-setup}

# First, let's use the summary dataset we created on the last slide
enroll_Summary_Stats_by_Stype <- raw_school_data %>% 
  filter(!is.na(enroll)) %>%
  group_by(stype) %>%
  summarize(
    Mean_enroll = mean(enroll),
    Max_enroll = max(enroll),
    Min_enroll = min(enroll),
   `Number of Schools` = n()
  )

# Make the mean bar chart, then use add_trace to add the other types of variables
min_bar_chart <- plot_ly(data =enroll_Summary_Stats_by_Stype,
                       type = "bar",
                       x = ~stype ,
                       y = ~`Min_enroll`,
                       name = "Min Enrollment")

min_bar_chart_with_mean <- min_bar_chart %>%
  add_trace(y = ~`Mean_enroll`,name = "Mean Enrollment")

min_bar_chart_with_mean_and_max <- min_bar_chart_with_mean %>%
  add_trace(y = ~`Max_enroll`,name = "Max Enrollment")


```

```{r D6, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# Look at the chart from the previous slide
min_bar_chart_with_mean_and_max

# Change the axis titles
bar_chart_new_layout <- min_bar_chart_with_mean_and_max %>%
  layout(yaxis = list(title = 'Number of Students'),
         xaxis = list(title = 'School Type'),
         title = "Enrollment by School Type")
  

bar_chart_new_layout
```

## Plotting Data - Colors 

The marker argument of the plot_ly or add_trace functions determines appearance 
This link has a lot more information about bar chart appearance: https://plotly.com/r/bar-charts/

```{r D7-setup}
enroll_Summary_Stats_by_Stype <- raw_school_data %>% 
  filter(!is.na(enroll)) %>%
  group_by(stype) %>%
  summarize(
    Mean_enroll = mean(enroll),
    Max_enroll = max(enroll),
    Min_enroll = min(enroll),
   `Number of Schools` = n()
  )

```


```{r D7, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# You can use the marker argument to change the bar's appearance
# Line describes the outline of the marker
mean_bar_chart <- plot_ly(data =enroll_Summary_Stats_by_Stype,
                       type = "bar",
                       x = ~stype ,
                       y = ~`Mean_enroll`,
                       name = "Mean Enrollment",
                       marker = list(color = 'lightblue',
                                     line = list(color = 'blue', width = 5)
                                     )
                       )

mean_bar_chart 

```

## Plotting Data - Labels 

The text and textposition arguments can be used to label the charts 

```{r D8-setup}
enroll_Summary_Stats_by_Stype <- raw_school_data %>% 
  filter(!is.na(enroll)) %>%
  group_by(stype) %>%
  summarize(
    Mean_enroll = mean(enroll),
    Max_enroll = max(enroll),
    Min_enroll = min(enroll),
   `Number of Schools` = n()
  )

```


```{r D8, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# We can label each column with its value
mean_bar_chart <- plot_ly(data =enroll_Summary_Stats_by_Stype,
                       type = "bar",
                       x = ~stype ,
                       y = ~`Mean_enroll`,
                       text = ~`Mean_enroll`,
                       textposition = "middle"
                       )

mean_bar_chart 

# We can even apply a function to the text to format it more legibly 
mean_bar_chart2 <- plot_ly(data =enroll_Summary_Stats_by_Stype,
                       type = "bar",
                       x = ~stype ,
                       y = ~`Mean_enroll`,
                       text = ~round(`Mean_enroll`),
                       textposition = "outside"
                       )

mean_bar_chart2


```

## Plotting Data - Scatterplots 

Changing the type argument will yield a different chart type in Plotly
The mode argument is either "markers" or "lines" and applies only to scatterplots

```{r D9, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

scatterplot <- plot_ly(data = raw_school_data,
                       type = "scatter",
                       x = ~meals,
                       y = ~api00,
                       mode="markers")

# Customize the scatterplot
scatterplot_formatted <- scatterplot %>% 
  layout(title = "Scatterplot of Students Eligible for Subsidized Meals vs API Scores",
         xaxis = list(title = "Percent of Students Eligible for Subsized Meals"),
         yaxis = list(title = "Scores 2000"))

scatterplot_formatted

```


## Plotting Data - Scatterplots 

Changing the type argument will yield a different chart type in Plotly
The mode argument is either "markers" or "lines" and applies only to scatterplots

```{r D9, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

scatterplot <- plot_ly(data = raw_school_data,
                       type = "scatter",
                       x = ~meals,
                       y = ~api00,
                       mode="markers")

# Customize the scatterplot
scatterplot_formatted <- scatterplot %>% 
  layout(title = "Scatterplot of Students Eligible for Subsidized Meals vs API Scores",
         xaxis = list(title = "Percent of Students Eligible for Subsized Meals"),
         yaxis = list(title = "Scores 2000"))

scatterplot_formatted

```


## Plotting Data - Histograms 

Changing the type to "histogram" will yield a histogram
There are so many plotly options for graphics, it's worth perusing the documentation:
https://plotly.com/r/ 

```{r D10, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# Create a histogram
histogram_api100 <- plot_ly(data = raw_school_data,
                       type = "histogram",
                       x = ~api00,
                       name="API Score 2000"
                       )

# Overlay two histograms on top of one another
histogram_comparison <- histogram_api100 %>%
  add_trace(type="histogram", x = ~api99,alpha = 0.6,name="API Score 1999") %>%
  layout(barmode="overlay")

histogram_comparison

```

## Exporting Graphics 

You can save your graphics as static images, or as interactive html files. 

```{r D11-setup}

# Create a histogram
histogram_api100 <- plot_ly(data = raw_school_data,
                       type = "histogram",
                       x = ~api00,
                       name="API Score 2000"
                       )

# Overlay two histograms on top of one another
histogram_comparison <- histogram_api100 %>%
  add_trace(type="histogram", x = ~api99,alpha = 0.6,name="API Score 1999") %>%
  layout(barmode="overlay")

```

```{r D11, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# Recall from the last slide we have a plotly object called histogram_comparison
class(histogram_comparison)

# Save as a static image: it is possible to do this with some dependencies
https://plotly-r.com/exporting-static-images.html#with-code

# You can also save directly from the Plotly chart itself using the camera button

# Save as HTML File
library(htmlwidgets)
#saveWidget(histogram_comparison, "histogram_plotly_chart.html")


```






