---
title: "R Workshop - Part 6 (Survey Data)"
author: "Peter Herman"
date: "2023-06-26"
output: slidy_presentation
runtime: shiny_prerendered
---

```{r setup, include=FALSE}

library(learnr)
library(haven)
library(srvyr)
library(survey)
library(dplyr)

setwd(r"()")

raw_survey_data <- read_sav("Example_Datasets/ATP W23.sav")


survey_data_recoded <- raw_survey_data %>%
  mutate(THERMO2_THERMTRUMP_W23_RECODED = case_when(
    THERMO2_THERMTRUMP_W23 == 999 ~ NA_real_,
    .default = as.numeric(THERMO2_THERMTRUMP_W23)
  ))

PEW_design_srvyr <- survey_data_recoded %>%
  as_survey_design(
    weights=WEIGHT_W23
  )

options(warn = -1)
options(datatable.verbose = FALSE)

```

## Reading in Data

Press K in your browser to avoid the site showing the next slide automatically.

This dataset comes from Pew Research Center’s American Trends Panel, and is a .sav file that exported from SPSS. The haven library allows us to import this file into R. There is also a function for reading in SAS .sas7bdat files. 
https://www.pewresearch.org/politics/dataset/american-trends-panel-wave-23/ 


```{r E1, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5, echo=FALSE,message=FALSE}

library(haven)
library(srvyr)
library(survey)
library(dplyr)

setwd(r"()")

raw_survey_data <- read_sav("Example_Datasets/ATP W23.sav")


```

## Recoding Survey Data

Respondents from the 2016 American Trends Panel were asked to rate Donald Trump on a scale of 1-100. This variable is called THERMO2_THERMTRUMP_W23.

Some of these respondents refused to answer the question. We should recode these respondents as NA.

Since we read this data in from a SPSS file, we can read the label on the variable, and that will tell us which values need to be recoded. 

```{r E2, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5, echo=FALSE,message=FALSE}

print_labels(raw_survey_data$THERMO2_THERMTRUMP_W23)

survey_data_recoded <- raw_survey_data %>%
  mutate(THERMO2_THERMTRUMP_W23_RECODED = case_when(
    THERMO2_THERMTRUMP_W23 == 999 ~ NA_real_,
    .default = as.numeric(THERMO2_THERMTRUMP_W23)
  ))

```

## Setting up a survey design object

We will be using the srvyr library, which allows for dplyr-like syntax around the functions from the popular survey library. 

To start- we need to define a survey design object, which is sort of like a dataframe but with the weight field specified.

```{r E3, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5,message=FALSE}

# There are many options available in the as_survey_design function, for strata, clusters, sampling probabilities etc

# However, we're just specifying weights
PEW_design_srvyr <- survey_data_recoded %>%
  as_survey_design(
    weights=WEIGHT_W23
  )

```

## Survey Summary Statistics 

We can use the dplyr summarize syntax along with the survey_mean function from the survey package to calculate the mean Trump rating among panel respondents.

Note that we must remove na's from our mean calculation using the na.rm argument

```{r E4, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5,message=FALSE}

PEW_design_srvyr %>%
  summarize(mean_trump_fav = survey_mean(THERMO2_THERMTRUMP_W23_RECODED,na.rm = TRUE)) %>%
  print()

# Note how the survey mean differs from the unweighted mean
survey_data_recoded  %>%
  summarize(mean_trump_fav = mean(THERMO2_THERMTRUMP_W23_RECODED,na.rm = TRUE)) %>%
  print()
  
```

## Survey Summary by Group 

We can group respondents by race/ethnicity (F_RACETHN_RECRUITMENT) using the dplyr group_by syntax

```{r E5, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5,message=FALSE}

PEW_design_srvyr %>%
  group_by(F_RACETHN_RECRUITMENT) %>%
  summarize(mean_trump_fav = survey_mean(THERMO2_THERMTRUMP_W23_RECODED,na.rm = TRUE)) %>%
  print()

```




