---
title: "R Workshop - Part 2 (Reading and Exploring Data)"
author: "Peter Herman"
date: "2023-06-26"
output: slidy_presentation
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(readr)
library(dplyr)

setwd(r"()")
raw_school_data <- read_delim("Example_Datasets/California_Academic_Performance_Index.txt",delim="\t")

```

## Reading in Data

Press K in your browser to avoid the site showing the next slide automatically.

We can use setwd to tell R where to look for our data
read_delim from the readr library should be used to read in data. 

```{r B1, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

library(readr)

# setwd (set working directory)
setwd(r"()")

# read_delim will tell you the data type of each column
raw_school_data <- read_delim("Example_Datasets/California_Academic_Performance_Index.txt",delim="\t")



```

## Seeing information about a function

If we want to know more about the read_delim function and what options are available, we can use the ? symbol

```{r B2, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

?read_delim

```

## Other ways of reading in data

R can handle most types of data, it's a just a matter of finding the correct library

```{r B3, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

library(haven)
?read_sas

library(readxl)
?read_xlsx

```

## Exploring a dataset - column names

We can use the names() function to get a vector of the column names of a data frame

```{r B4, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

names(raw_school_data)


```

## Exploring a dataset - dimensions

We can use the dim() function to see how many rows and columns there are in the dataset

```{r B5, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

dim(raw_school_data)


```

## Exploring a dataset - head

We can look at the first five rows of our data frame using head

```{r B6, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

head(raw_school_data)


```

## Exploring a dataset - View

We can see the full dataset using View

```{r B7, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

View(raw_school_data)

```

## Exploring a dataset - summary

You can get a sense of the distribution of the data in each column using summary
See the data dictionary here: https://r-survey.r-forge.r-project.org/survey/html/api.html

```{r B8, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# Take a look at the average class size for years K-3
summary(raw_school_data$acs.k3)

# There are 1823 NA's, as our school level data frame contains high, middle, and elementary schools

# Rather than use the $ operator, we can also use pull() from the dplyr function, which does the same thing.
library(dplyr)

# Take a look at the average class size for years 4-6
raw_school_data %>%
  pull(acs.46) %>%
  summary()


```

## Exploring a dataset - table

We can create quick frequency tables in R using the table() function

```{r B9, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# See how many schools of each type we have
table(raw_school_data$stype)

# By default, table will not include missing data
# Here's a frequency of class sizes for years K-3
table(raw_school_data$acs.k3)

# Here's a frequency of class sizes including missing data
table(raw_school_data$acs.k3,useNA = c("always"))

```

