---
title: "R Workshop - Part 1 (Data Types and Functions)"
author: "Peter Herman"
date: "2023-06-26"
output: slidy_presentation
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(readr)
library(dplyr)


```

# The Very Basics of Using R

## Working in the R Console

Press K in your browser to avoid the site showing the next slide automatically.

When you enter code into the R console, you get a little [1] next to the output. Ignore it! 

```{r A1, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

1 + 2

```

## Variable Assignment using <-

In R, we can create variables using the <- operator

```{r A2, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

my_variable <- 2

2 * my_variable

```

Unfortunately, R is case sensitive. Trying to use the same variable in all caps does not work

```{r A3, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

3 * MY_VARIABLE

```

## Data Classes

There are five data classes in R, we can use typeof or class to see what type of variable we have

```{r A4, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# Characters
my_character_variable <- "b"

# Integers (rarely used)
my_intiger <- 2L

# Numbers
my_numeric_variable <- 2.6

# Logical (either TRUE or FALSE)
my_logical_variable <- TRUE

# For now, let's forget about the fifth data type, Date

typeof(my_character_variable)

```

## Data Structures

The most common data structures in R are vectors, data frames, and lists

```{r A5, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}


# Vectors are a collection of variables of the same type
# the c stands for "combine"
my_numeric_vector <- c(14,54,63)


# You can use the square brackets to get an individual vector value
print(my_numeric_vector[1])
print(my_numeric_vector[2])

# operations applied to vector apply to all the elements in a vector
print(my_numeric_vector*2)

# You can use %in% to see if an element is in a vector
print(23 %in% my_numeric_vector)
print(14 %in% my_numeric_vector)

```

## Data Frames

Data frames are rectangular data in R. 

```{r A6, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# First, define two vectors
my_numeric_vector <- c(14,54,63)
my_character_vector <- c("Value 1", "Value 2","Value 3")

# You can use these vectors to become columns in a data frame
my_data_frame <- data.frame(
  Labels = my_character_vector,
  Values = my_numeric_vector
)

print(my_data_frame)

# You can use the $ operator to get at an individual column in a data frame
print(my_data_frame$Values)

```

## Lists

Lists are like vectors, but they can hold data of different types

```{r A7, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

my_list <- list(14,"orange",1.9)

# You can get individual list elements using [[]]
print(my_list[[1]])
print(my_list[[2]])

# You can add data to an empty slot of a list
my_list[[4]] <- "red"
print(my_list)

```

## Functions

In R we can create functions that take some input, called arguments and return some output 

```{r A8, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# Define a function using the function keyword
# our argument in this function is called "number"
multiply_by_two <- function(number){
  number_multiplied <- number * 2
  return(number_multiplied)
}

# We can include the name of the argument in the function call, but do not have to 
print(multiply_by_two(number=20))
print(multiply_by_two(4))

# The variables within the function are not stored in memory
print(number_multiplied)

```

## The Pipe Operator

In R there are two ways of calling functions

```{r A9, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# First, define the functions
multiply_by_two <-function(num){ return(num*2)}

# Typical, base R way to call a function
multiply_by_two(80)

# Using the dplyr library, we can use the %>% operator
library(dplyr)

80 %>% multiply_by_two()

```

## The Pipe Operator in functions with two arguments

When using the pipe operator, only the first argument to each function is 

```{r A10, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

add_these_numbers <- function(number1,number2){
  my_added_output <- number1 + number2
  return(my_added_output)
}

# Not using the pipe
add_these_numbers(5,10)

# Using the pipe
5 %>% add_these_numbers(10)


```


## Chaining Arguments

The use case for the pipe is chaining multiple functions together

```{r A11, exercise=TRUE, exercise.eval=FALSE, fig.height=5.5}

# First define the functions
multiply_by_two <-function(num){ return(num*2)}
add_these_numbers <- function(number1,number2){ return(number1 + number2)}

# We can call the pipe operator twice, chaining the functions together
My_New_Number <- 4 %>%
  multiply_by_two() %>%
  add_these_numbers(2)

print(My_New_Number)

# If we wrote this in the base R way, there would be a lot of parenthesis 
My_New_Number <- add_these_numbers(multiply_by_two(4),2)

```





